---
title: "MxMonks Final"
author: "Liz Hobson"
date: "July 17, 2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries
```{r, echo=FALSE}
library(scales)
library(zoo)
library(plotrix)
library(maps)
library(mapproj)
library(maptools)
library(sp)
library(reshape2)
library(lubridate)
library(plyr)
library(dplyr)
```


## Import parsed trade & citizen science data
Set location for getting data
```{r, echo=TRUE}
LOC.finaldata <- "C:/Users/ehobs/Dropbox/P-monk MX invasion/MX-Monks/PLOSONE_FINAL_DATA/"

#NEW CITES DATA:CITES-new28Feb
cites <- read.csv(file=paste(LOC.finaldata, "cites_imports.csv", sep=""))

dgvs <- read.csv(file=paste(LOC.finaldata, "dgvs_imports.csv", sep=""))

citsci <- read.csv(file=paste(LOC.finaldata, "citizen_science_reports.csv", sep=""))

# add dummy shipment counter to DGVS data
dgvs$shipment.counter <- 1
dgvs$shipment.counter[dgvs$year==2015] <- 0
# check

str(cites)
str(dgvs)
str(citsci)

```

## SUMMARIZE CITES DATA 
(already subset, so purpose = live, Term = "T" for commercial)

```{r, echo=TRUE}
# summarize number imported indivs X reporter type per year, per purpose
grouped.MX <- group_by(cites, Year)
cites.summXyear <- as.data.frame(summarise(grouped.MX, 
                                            CITESindivsXIMreps=sum(Importer.reported.quantity, na.rm=TRUE),
                                            CITESindivsXEXreps=sum(Exporter.reported.quantity, na.rm=TRUE)
                                            )) 

cites.summXyear <- arrange(cites.summXyear, Year)
str(cites.summXyear)

cites.summXyear$CITESmin.rep <- pmin(cites.summXyear$CITESindivsXIMreps, cites.summXyear$CITESindivsXEXreps) # return minimum reported import quantity
cites.summXyear$CITESmax.rep <- pmax(cites.summXyear$CITESindivsXIMreps, cites.summXyear$CITESindivsXEXreps) # return maximum reported import quantity
cites.summXyear$CITESdiff.reps <- cites.summXyear$CITESmax.rep - cites.summXyear$CITESmin.rep # return difference between maximum and minimum reported import quantity

```

# set up blank dataframe to fill with answers to summary questions
Want historical perspective on CITES data that we can't get from DGVS

```{r, echo=FALSE}
summary<- data.frame(question=character(),
                      answer=as.numeric(),
                      stringsAsFactors=FALSE) 
```

```{r, echo=TRUE}
q3 <- cbind("q3", "what is max difference between importer and exporter reports?","CITES",
      max(cites.summXyear$CITESdiff.reps))

q3b <- cbind("q3b", "what is total diff btw importer and exporter report #s?","CITES",
      sum(cites.summXyear$CITESindivsXIMreps) - sum(cites.summXyear$CITESindivsXEXreps))


q1 <- cbind("q1", "#individuals total were imported for whole dataset? (IMrep)","CITES",
sum(cites$Importer.reported.quantity, na.rm=TRUE))

q2 <- cbind("q2", "#individuals total were imported for whole dataset? (EXrep)","CITES",
sum(cites$Exporter.reported.quantity, na.rm=TRUE))

q4 <- cbind("q4", "what is difference between importer reports & exporter reports in # indivs? (from raw df)","CITES",
            sum(cites$Importer.reported.quantity, na.rm=TRUE) - sum(cites$Exporter.reported.quantity, na.rm=TRUE))

q5 <- cbind("q5", "what is difference between IM&EX reports in # indivs? (summ df)","CITES",
sum(cites.summXyear$CITESindivsXIMreps) - sum(cites.summXyear$CITESindivsXEXreps)) #check, this and q4 should be same

q6a <- cbind("q6a", "imported reports are xx% of exported reports? (raw df)","CITES",
  round(sum(cites$Importer.reported.quantity, na.rm=TRUE) / sum(cites$Exporter.reported.quantity, na.rm=TRUE) *100,0))

q6b <- cbind("q6b", "imported reports are xx% of exported reports? (fr summ df)","CITES",
round((sum(cites.summXyear$CITESindivsXIMreps) / sum(cites.summXyear$CITESindivsXEXreps)) *100, 0))

q7 <- cbind("q7", "what year (1975-2015) did first imports occur?","CITES",
min(cites$Year, na.rm=TRUE))

q8 <- cbind("q8", "what year (1975-2015) did last imports occur?","CITES",
max(cites$Year, na.rm=TRUE))

# find earlier imports, before 2000 (ie to end of 1999)
cites.b42000 <- subset(cites, Year<2000)

q9 <- cbind("q9", "#indivs commercially imported b4 2000 (ie by end of 1999? (IM.reported)","CITES",
sum(cites.b42000$Importer.reported.quantity, na.rm=TRUE))

q10 <- cbind("q10", "#indivs commercially imported b4 2000 (ie by end of 1999? (EX.reported)","CITES",
sum(cites.b42000$Exporter.reported.quantity, na.rm=TRUE))

q10a <- cbind("q10a", "%indivs commercially imported b4 2000 (ie by end of 1999? (IN.reported)","CITES",
round(
  (sum(cites.b42000$Importer.reported.quantity, na.rm=TRUE)/sum(cites.summXyear$CITESindivsXIMreps))*100,2)
)

q10b <- cbind("q10b", "%indivs commercially imported b4 2000 (ie by end of 1999? (EX.reported)","CITES",
round(
  (sum(cites.b42000$Exporter.reported.quantity, na.rm=TRUE)/sum(cites.summXyear$CITESindivsXEXreps))*100,2)
)

q10c <- cbind("q10c", "what % of indivs imported btw start2008-end 2015? (IM.rep)","CITES",
round((sum(cites.summXyear$CITESindivsXIMreps[cites.summXyear$Year>2007]) / sum(cites.summXyear$CITESindivsXIMreps)) *100,0))

q11 <- cbind("q11", "what % of indivs imported between 2008-end 2015? (EX.rep)","CITES",
round((sum(cites.summXyear$CITESindivsXEXreps[cites.summXyear$Year>2007]) / sum(cites.summXyear$CITESindivsXEXreps)) *100,0))

q11a <- cbind("q11a", "how many indivs imported 2000-end 2015? (IM.rep)","CITES",
sum(cites.summXyear$CITESindivsXIMreps[cites.summXyear$Year>1999]))

q11b <- cbind("q11b", "how many indivs imported 2000-end 2015? (EX.rep)","CITES",
sum(cites.summXyear$CITESindivsXEXreps[cites.summXyear$Year>1999]))
```

### get total # individuals from each exporting country to MX for live commercial exports
```{r, echo=TRUE}
grouped.origins <- group_by(cites, Year, Exporter)
cites.summXcountryXyear <- as.data.frame(summarise(grouped.origins,  
                                            nindivsXimporterreps=sum(Importer.reported.quantity, na.rm=TRUE), 
                                            nindivsXexporterreps=sum(Exporter.reported.quantity, na.rm=TRUE)
                                            ))
```

### of TOTAL historical exports (commercial/live to MX, CITES), what % individuals came from each exporting country?
```{r, echo=TRUE}
grouped.origins2 <- group_by(cites, Exporter)
cites.summXcountry <- as.data.frame(summarise(grouped.origins2,  
                                            nindivsXimporterreps=sum(Importer.reported.quantity, na.rm=TRUE), 
                                            nindivsXexporterreps=sum(Exporter.reported.quantity, na.rm=TRUE)
                                            ))

cites.summXcountry$percent.indivs.exportedIMreps <- round((cites.summXcountry$nindivsXimporterreps / sum(cites.summXcountry$nindivsXimporterreps)*100),0)
cites.summXcountry$percent.indivs.exportedEXreps <- round((cites.summXcountry$nindivsXexporterreps / sum(cites.summXcountry$nindivsXexporterreps)*100),0)
as.data.frame(t(cites.summXcountry))
```
             
### how many unique countries were commercial/live exporters to MX?
```{r}
unique(cites$Exporter)
length(unique(cites$Exporter)) 
```

### More import questions
```{r}
str(cites.summXcountry)
q12 <- cbind("q12", "#indivs that originated from Uruguay for whole history? (IM.rep)", "CITES",
cites.summXcountry$nindivsXimporterreps[cites.summXcountry$Exporter=="UY"])

q13 <- cbind("q13", "#indivs that originated from Uruguay for whole history? (EX.rep)", "CITES",
cites.summXcountry$nindivsXexporterreps[cites.summXcountry$Exporter=="UY"])

q14 <- cbind("q14", "diff btw IM&EX #indivs from Uruguay whole history?", "CITES",
cites.summXcountry$nindivsXimporterreps[cites.summXcountry$Exporter=="UY"] - cites.summXcountry$nindivsXexporterreps[cites.summXcountry$Exporter=="UY"])

q15 <- cbind("q15", "diff % IMvsEX #indivs from Uruguay whole history?", "CITES",
round((cites.summXcountry$nindivsXimporterreps[cites.summXcountry$Exporter=="UY"] / cites.summXcountry$nindivsXexporterreps[cites.summXcountry$Exporter=="UY"])*100,0)
)
```


# SUMMARIZE DGVS DATA

### summary questions: number individuals imported (data from DGVS)
```{r}
q16 <- cbind("q16", "#individuals total commercially imported (whole dataset)?", "DGVS",
sum(dgvs$quantity))

q17 <- cbind("q17", "between which years? start", "DGVS",
min(dgvs$year))

q18 <- cbind("q18", "between which years? end", "DGVS",
max(dgvs$year))

#subset, find data start 2008 to end 2015
DGVS.2008to2015 <- subset(dgvs, year>2007)

q19 <- cbind("q19", "how many individuals total were imported beginning 2008 to end 2015?", "DGVS",
sum(DGVS.2008to2015$quantity))

q20 <- cbind("q20", "what % of total imported individuals were imported between 2008 and end of 2014?", "DGVS",
round((sum(DGVS.2008to2015$quantity) / sum(dgvs$quantity)) *100,0))
```

## summarize # shipments imported
```{r}
q22 <- cbind("q22", "how many commercial shipments total were imported 1999 to 2015?", "DGVS",
sum(dgvs$shipment.counter))

q23 <- cbind("q23", "how many shipments total were imported beginning 2008 to end of 2015?", "DGVS",
sum(DGVS.2008to2015$shipment.counter))

q24 <- cbind("q24", "what % of total imported individuals were imported between 2008 and end of 2014?", "DGVS",
round((sum(DGVS.2008to2015$shipment.counter) / sum(dgvs$shipment.counter)) *100,0))

```

## summarize # indivs X year, # shipments X year
```{r, echo=TRUE}
str(DGVS.2008to2015)
str(dgvs)

# get total # individuals and # shipments from each exporting country to MX for live commercial exports
G.origins <- group_by(dgvs, year, origin.country)
DGVS.shipsXcountryXyr <- as.data.frame(summarise(G.origins, total.shipments=sum(shipment.counter), total.indivs=sum(quantity)))


q25 <- cbind("q25", "how many unique countries were commercial/live exporters to MX? (country of origin) (2000-2015)", "DGVS",
length(unique(DGVS.shipsXcountryXyr$origin.country)))

unique(DGVS.shipsXcountryXyr$origin.country)
str(DGVS.shipsXcountryXyr)
```

### of TOTAL DGVS-reported imports (commercial/live to MX), what % individuals came from each exporting country (2000-2015)?
```{r, echo=TRUE}
grouped.exports.DGVS <- group_by(DGVS.shipsXcountryXyr, origin.country)
DGVS.shipsXcountry <- as.data.frame(summarise(grouped.exports.DGVS, total.shipments=sum(total.shipments), total.indivs=sum(total.indivs)))
DGVS.shipsXcountry$percent.indivs.exported <- round((DGVS.shipsXcountry$total.indivs / sum(DGVS.shipsXcountry$total.indivs)*100),0)
DGVS.shipsXcountry$percent.ships.exported <- round((DGVS.shipsXcountry$total.shipments / sum(DGVS.shipsXcountry$total.shipments)*100),0)
str(DGVS.shipsXcountry)
```

### of TOTAL start 2008- end 2015 exports (commercial/live to MX), what % individuals came from each exporting country?

```{r, echo=TRUE}
DGVS.export.countries.2008to2015 <- subset(DGVS.shipsXcountryXyr, year>2007 & total.shipments>0) #exclude the dummy "0" shipment without origin country
str(DGVS.export.countries.2008to2015)
Gr.exports.DGVS.2008to2015 <- group_by(DGVS.export.countries.2008to2015, origin.country)
DGVS.export.countries.2008to2015.summ <- as.data.frame(summarise(Gr.exports.DGVS.2008to2015, total.shipments=sum(total.shipments), total.indivs=sum(total.indivs)))

DGVS.export.countries.2008to2015.summ$percent.indivs.exported <- round((DGVS.export.countries.2008to2015.summ$total.indivs / sum(DGVS.export.countries.2008to2015.summ$total.indivs)*100),0)

DGVS.export.countries.2008to2015.summ$percent.ships.exported <- round((DGVS.export.countries.2008to2015.summ$total.shipments / sum(DGVS.export.countries.2008to2015.summ$total.shipments)*100),0)

str(DGVS.export.countries.2008to2015.summ)
```

### get total # individuals and # shipments from across all exporting countrys to MX for live commercial exports

```{r, echo=TRUE}
G.nindivsXyear <- group_by(dgvs, year)
DGVS.nindivsXyear <- as.data.frame(summarise(G.nindivsXyear, total.DGVS.shipments=sum(shipment.counter), total.DGVS.total.indivs=sum(quantity)))
str(DGVS.nindivsXyear)
```

## Compile trade data DGVS + CITES

```{r, echo=TRUE}
str(cites.summXyear)
str(DGVS.nindivsXyear) 

TRADExYR.compiled <- merge(cites.summXyear, DGVS.nindivsXyear, by.x="Year", by.y="year", all=TRUE, fill=0)
#TRADExYR.compiled[is.na(TRADExYR.compiled)] <- 0
str(TRADExYR.compiled)

TRADExYR.compiled$CITESindivsXIMreps[is.na(TRADExYR.compiled$CITESindivsXIMreps)] <- 0
TRADExYR.compiled$CITESindivsXEXreps[is.na(TRADExYR.compiled$CITESindivsXEXreps)] <- 0
TRADExYR.compiled$CITESmin.rep[is.na(TRADExYR.compiled$CITESmin.rep)] <- 0
TRADExYR.compiled$CITESmax.rep[is.na(TRADExYR.compiled$CITESmax.rep)] <- 0
TRADExYR.compiled$CITESdiff.reps[is.na(TRADExYR.compiled$CITESdiff.reps)] <- 0

TRADExYR.compiled$diffimportsDGVSxCITES.IM <- TRADExYR.compiled$total.DGVS.total.indivs - TRADExYR.compiled$CITESindivsXIMreps
TRADExYR.compiled$which.higher <- ifelse(TRADExYR.compiled$total.DGVS.total.indivs > TRADExYR.compiled$CITESindivsXIMreps, "DGVS higher", ifelse(TRADExYR.compiled$total.DGVS.total.indivs==TRADExYR.compiled$CITESindivsXIMreps, "same", ifelse(TRADExYR.compiled$total.DGVS.total.indivs<TRADExYR.compiled$CITESindivsXIMreps,"lower", "no data")))

TRADExYR.compiled$which.higher[is.na(TRADExYR.compiled$which.higher)] <- "NO DATA"

str(TRADExYR.compiled)


#subset X year
TRADExYR.compiled.1999toend2015 <- subset(TRADExYR.compiled, Year>1998 & Year <2016)
TRADExYR.compiled.2000toend2015 <- subset(TRADExYR.compiled, Year>1999 & Year <2016)
TRADExYR.compiled.1999toend2014 <- subset(TRADExYR.compiled.1999toend2015, Year<2015)
```

# compare DGVS with CITES data
```{r}
q26 <- cbind("q26", "how many years did CITES imported-reported number match DGVS?", "DGVS v CITES.IM",
yrs.DGVS.same <- length(TRADExYR.compiled.2000toend2015$which.higher[TRADExYR.compiled.2000toend2015$which.higher=="same"]))
yrs.DGVS.same #number of years of match
total.yrs <- 2015-2000

q27 <- cbind("q27", "what % years did CITES imported-reported number match DGVS?", "DGVS v CITES.IM",
round(yrs.DGVS.same / total.yrs *100, 0))

q28 <- cbind("q28", "total imported individuals DGVS reported start 2000 to end 2015?", "DGVS",
total.DGVS <- sum(TRADExYR.compiled.2000toend2015$total.DGVS.total.indivs))
total.DGVS

q28 <- cbind("q28", "total imported individuals CITES.IM reported start 2000 to end 2015?", "CITES.IM",
total.CITES.IM <- sum(TRADExYR.compiled.2000toend2015$CITESindivsXIMreps))
total.CITES.IM

q29 <- cbind("q29", "what is total discrepancy: CITES-importer minus DGVS reported?", "DGVS v CITES.IM",
total.CITES.IM - total.DGVS)

q29a <- cbind("q29a", "what is total discrepancy: CITES-exporter(2000-2015) minus DGVS reported?", "DGVS v CITES.IM",
total.DGVS - sum(TRADExYR.compiled.2000toend2015$CITESindivsXEXreps))

q30 <- cbind("q30", "how many DGVS shipments occured start 2000 to end 2015?", "DGVS",
total.DGVS.shipments <- sum(TRADExYR.compiled.2000toend2015$total.shipments.DGVS))
total.DGVS.shipments
```

## Compile summary of answers to trade questions
```{r, echo=TRUE}
summary <- as.data.frame(rbind(q1,q2,q3,q3b,q4,q5,q6a,q6b,q7,q8,q9,q10,q10a,q10b,q10c,q11,q11a,q11b,q13,q14,q15,q16,q17,q18,q19,q20,q22,q23,q24,q25,q26,q27,q28,q29, q30)) #exclude 12 and 21

colnames(summary) <- c("Qid", "question", "data", "answer")
summary

```

# SUMMARIZE CITIZEN SCIENCE REPORTS
```{r, echo=TRUE}
str(citsci)
str(citsci)

#set date as.Date
citsci$date <- as.Date(citsci$date)
citsci$year <- year(citsci$date)
citsci$report.dummy <- 1

############## Basic stats on full DB (good date/loc, pre-2016)
min(citsci$date) # what was date of first report in cit sci database?
max(citsci$date) # what was date of last report in cit sci database?
length(citsci$date) # how many reports were filed?
length(unique(citsci$date)) # how many unique dates had reports?

citsci[which.min(citsci$date),] #extract earliest record

# get list of all unique cities where monks observed before 2016 
#(check city/state combos to make sure there aren't more "La Paz"s hiding, ie duplicate city names)
list.unique.citiesstates <- unique(citsci[c("city", "state")])
list.unique.cities <- unique(citsci$city)
list.unique.states <- unique(citsci$state)

#total number unique cities (should be same for both calls, and =93)
length(list.unique.citiesstates$city)
length(list.unique.cities)

# get list of average location of obs X city using [DATA.trim] & [list.unique.citiesstates]

G.locs <- group_by(citsci, city, state)
avg.city.loc.MASTER <- as.data.frame(summarise(G.locs, avg.cityLAT=mean(latitude), avg.cityLOG=mean(longitude)))
str(avg.city.loc.MASTER)


#find # unique cities X year
citsci.unique.citiesXyr <- unique(citsci[c("year", "city")])
citsci.unique.citiesXyr$dummy.unique.city <- 1
unique(is.na(citsci.unique.citiesXyr$city)) # should all be "FALSE" = no missing cities


#how many unique cities had reports per year?
grouped.citys <- group_by(citsci.unique.citiesXyr, year)
citsci.unique.citiesXyr.summ <- as.data.frame(summarise(grouped.citys, unique.cities=sum(dummy.unique.city)))
citsci.unique.citiesXyr.summ

#using dplyr, group obs X yr
grouped <- group_by(citsci, year)
citsci.reportsXyr <- as.data.frame(summarise(grouped, total.reports=sum(report.dummy)))
citsci.reportsXyr

#add unique reports X yr to unique cities X yr
citsciXYRxcityxreports <- cbind(citsci.reportsXyr, citsci.unique.citiesXyr.summ$unique.cities)
str(citsciXYRxcityxreports)
names(citsciXYRxcityxreports) <- c("year", "n.citsci.reports", "n.citsci.cities")
```

## add region data 

```{r, echo=TRUE}
mx.states.atts <- read.table(file=paste(LOC.finaldata, "mx.states.attributes.txt", sep=""), stringsAsFactors = FALSE, sep = "\t", as.is=T,header=T)
str(mx.states.atts)

#import MX state names (with and without accents)
state.names <- read.csv("state.names.csv", sep = ",", as.is=T,header=T)
str(state.names)
#import region MASTER
reg7.master <- read.csv(paste(LOC.finaldata, "reg7.master.csv", sep=""))
str(reg7.master)

#merge state names with color X region
statesXregWcols <- merge(mx.states.atts, reg7.master, by="reg7.cat", all.x=T)
str(statesXregWcols)

statesXregWcols <- merge(statesXregWcols, state.names, by=c("state", "reg7.cat"))
str(statesXregWcols)


#merge state data with citizen science reports to summarize by region
DATA.reg7 <- merge(citsci, statesXregWcols, by="state", all.x=T)

str(DATA.reg7)

unique(is.na(DATA.reg7$reg7.cat)) #should all be false, no missing data 
```

## get # reports X city X year

```{r, echo=TRUE}
#how many citsci reports are there per year per NEW 7 region syst & city?
grouped.regionXcityXyr <- group_by(DATA.reg7, year, reg7.cat, city, state, reg.order)
reportsXyrXregionXcity <- as.data.frame(summarise(grouped.regionXcityXyr, total.obsXcityXyr=sum(report.dummy)))
str(reportsXyrXregionXcity)
reportsXyrXregionXcity$dummy.XcityreportingXyr <- 1 #adds dummy to count # cities reporting PER YEAR
```

## how many monk cities are in each new region (7 regs)? NEW: FLIP DATA AROUND, region, state, city X years
```{r, echo=TRUE}
CAST.citiesXyrs <- dcast(data=reportsXyrXregionXcity, formula=reg7.cat+reg.order+state+city~year, value.var="total.obsXcityXyr", fill=0)
CAST.citiesXyrs <- arrange(CAST.citiesXyrs, reg.order, state, city)
```


### now melt 
```{r, echo=TRUE}
str(CAST.citiesXyrs)
#melt(airquality, id.vars = c("month", "day"))
MELT.citiesXyrs <- melt(data=select(CAST.citiesXyrs, -reg7.cat, -reg.order), id.vars= c("state","city"), variable.name="year", value.name="num.reportsXcity")
str(MELT.citiesXyrs)


MELT.citiesXyrs.no0s <- subset(MELT.citiesXyrs, num.reportsXcity>0)
str(MELT.citiesXyrs.no0s)
MELT.citiesXyrs.no0s$year <- as.numeric(as.character(MELT.citiesXyrs.no0s$year))
MELT.citiesXyrs.no0s <- arrange(MELT.citiesXyrs.no0s, state, city)
str(MELT.citiesXyrs.no0s)
```

## add new citation data for dotplot
```{r, echo=TRUE}
#import data
citeMASTER <- read.csv(file=paste(LOC.finaldata, "LIT_citationMASTER.csv", sep=""))
cite.locsXyrs <- read.csv(file=paste(LOC.finaldata, "LIT_locationsXyears_citations_TRIMMED.csv", sep=""))
str(cite.locsXyrs)

#add citation field to merge by
cite.locsXyrs$citation <- paste(cite.locsXyrs$source, cite.locsXyrs$year.published, sep=" ")
```

## merge id.code for cites with location data, then add need new y axis in cases where cities are in citation data, not in citsci data
```{r, echo=TRUE}
locXyrs.citations <- merge(select(citeMASTER, cite.id, citation), cite.locsXyrs, by="citation")
locXyrs.citations <- arrange(locXyrs.citations, cite.id, year.sighted)

#add y.axis for cities! -- 
str(locXyrs.citations)

citations.cityXstate <- unique(locXyrs.citations[c("city", "state")])
str(citations.cityXstate)

citsci.cities <- unique(CAST.citiesXyrs[c("city", "state")])
str(citsci.cities)

#get unique city / state across citations and citizen science DBs
pre.MASTER.cities.citsciXcitations <- rbind(citsci.cities, citations.cityXstate)

MASTER.cities.CSciXcites <- unique(pre.MASTER.cities.citsciXcitations[c("city", "state")])
str(MASTER.cities.CSciXcites) # this is full list of all unique cities in either citsci or citation lists
```

## add regions back in (by merging on state)
```{r, echo=TRUE}
MASTER.cities.CSciXcites.Wstates.Wreg7s <- merge(MASTER.cities.CSciXcites, statesXregWcols, by="state", all.x=TRUE)
str(MASTER.cities.CSciXcites.Wstates.Wreg7s)

#now set new order
MASTER.cities.CSciXcites.Wstates.Wreg7s <- arrange(MASTER.cities.CSciXcites.Wstates.Wreg7s, reg.order, state, city)

#now add new y-axis coords
MASTER.cities.CSciXcites.Wstates.Wreg7s$yPlot.ord.CSciXcites <- seq(1, length(MASTER.cities.CSciXcites.Wstates.Wreg7s$city))
str(MASTER.cities.CSciXcites.Wstates.Wreg7s)

#now add y-axis coords back to citation data (to control where to plot text ids)
locXyrs.citations.wCOORDS <- merge(locXyrs.citations, MASTER.cities.CSciXcites.Wstates.Wreg7s, 
                                   by=c("city", "state"), all.x=TRUE)

locXyrs.citations.wCOORDS <- arrange(locXyrs.citations.wCOORDS, yPlot.ord.CSciXcites, year.sighted) #sort

```

# FIGURE 1: importation and sightings

## process data (combine shipping and citizen science reports)
```{r, echo=TRUE}
# merge shipping and citsci reports
str(citsci.reportsXyr)
str(TRADExYR.compiled)

trade.citscixyr.COMBO <- merge(TRADExYR.compiled, citsci.reportsXyr, by.x="Year", by.y="year", all.x=TRUE)
str(trade.citscixyr.COMBO)

#subset X year
trade.citscixyr.COMBO.1999toend2015 <- subset(trade.citscixyr.COMBO, Year>1998 & Year <2016)
str(trade.citscixyr.COMBO.1999toend2015)

trade.citscixyr.COMBO.1999toend2014 <- subset(trade.citscixyr.COMBO, Year<2015)
str(trade.citscixyr.COMBO.1999toend2014)
```

## PLOT FIGURE 1
```{r, echo=TRUE}
plot.new()
############## set up plot:
#setwd("C:/Users/ehobs/Dropbox/P-monk MX invasion/MX-Monks/PLOTS/")

###### save plot as SVG to mod in inkscape
#svg("SVGfig.shipsXsightings.LaPaz.fixed.ships.fixed.svg", width=7, height=7)

###################################### shipping plot CITES range of import reports vs export reports, now plot as range rectangles? (plot on bottom)
par(mfrow=c(2,1), mar=c(5,8,2,2))
with(trade.citscixyr.COMBO.1999toend2015, plot(Year, total.reports, las=2, xlim=c(1999, 2015), ylim=c(0,150000), ylab=NA, type="n", axes=F ))
# add lines for change in regs/bird flu

# add axes
axis(side = 1, at = seq(min(trade.citscixyr.COMBO.1999toend2015$Year), max(trade.citscixyr.COMBO.1999toend2015$Year), by=1), las=2)
axis(side = 2, las=2)
box()

abline(v=2008, lty=1, lwd=2, col=alpha("black", 0.2), xpd=F) 
abline(v=2014, lty=1, lwd=2, col=alpha("black", 0.2), xpd=F)

#underplot rectangles
rect(xleft=trade.citscixyr.COMBO.1999toend2015$Year-0.25, ybottom=trade.citscixyr.COMBO.1999toend2015$CITESmin.rep, 
     xright=trade.citscixyr.COMBO.1999toend2015$Year+0.25, ytop=trade.citscixyr.COMBO.1999toend2015$CITESmax.rep,
     col = "grey", border = alpha("black", 0.3), lwd=2)

#################################### overplot DGVS data

# SHIPPING PLOT [Rcode-shipments] DGVS data (X shipments)

#overplot data
lines(trade.citscixyr.COMBO.1999toend2015$Year, trade.citscixyr.COMBO.1999toend2015$total.DGVS.total.indivs, type="l", lwd=1.5, xpd=F)
points(trade.citscixyr.COMBO.1999toend2014$Year, trade.citscixyr.COMBO.1999toend2014$total.DGVS.total.indivs, las=2, 
       ylab=NA, cex=0.8, 
       pch=21, col="black", bg="black", xpd=F)


title(ylab = "Total individuals", cex.lab = 1, line = 4.5)
text(x=trade.citscixyr.COMBO.1999toend2015$Year, y=trade.citscixyr.COMBO.1999toend2015$total.DGVS.total.indivs+10000, 
     trade.citscixyr.COMBO.1999toend2015$total.DGVS.shipments, cex=0.75, srt=0, adj=0.5)

text(x=1999, y=140000, "Text labels = Number of shipments", cex=0.85, srt=0, adj=0)
mtext("(a)", side = 3, line = 0, outer = FALSE, cex = 1)


# SIGHTINGS PLOT [Rcode-sightings(newEbirddata)]

#str(reportsXyr)
#plot points, no axes
plot(trade.citscixyr.COMBO.1999toend2015$Year, trade.citscixyr.COMBO.1999toend2015$total.reports, type="o", las=3, lwd=1.5, pch=23, col="black", bg="grey", cex=1.0, axes = FALSE, xlab="Year", ylab=NA, xlim=c(min(trade.citscixyr.COMBO.1999toend2015$Year), max(trade.citscixyr.COMBO.1999toend2015$Year)), ylim=c(0,1300))

#add vertical line markers
abline(v=2008, lty=1, lwd=2, col=alpha("black", 0.2), xpd=F) 
abline(v=2014, lty=1, lwd=2, col=alpha("black", 0.2), xpd=F)
# add axes
axis(side = 1, at = seq(min(trade.citscixyr.COMBO.1999toend2015$Year), max(trade.citscixyr.COMBO.1999toend2015$Year), by=1), las=2)
axis(side = 2, las=2)
title(ylab = "Number of sightings", cex.lab = 1, line = 4.5)
box()

#add text labels for # unique cities (unique.cities)
text(x=trade.citscixyr.COMBO.1999toend2015$Year, y=trade.citscixyr.COMBO.1999toend2015$total.reports+100, 
     trade.citscixyr.COMBO.1999toend2015$total.reports, cex=0.75, srt=0, adj=0.5)
text(x=1999, y=1200, "Text labels = Number of cities with reports", cex=0.85, srt=0, adj=0)

mtext("(b)", side = 3, line = 0, outer = FALSE, cex = 1)
#dev.off()
```

# FIGURE 2: Map of invaded cities in Mexico
tutorial: https://www.students.ncl.ac.uk/keith.newman/r/maps-in-r-using-gadm

download shapes here: http://www.gadm.org/download -- level 1

## Import data
```{r, echo=TRUE}

mx.shapes <- readRDS(file=paste(LOC.finaldata, "MEX_adm1.rds", sep=""))

str(statesXregWcols)

all.states <- arrange(statesXregWcols, names.w.accents)
str(all.states)


lit.cities <- read.csv(file=paste(LOC.finaldata, "cities.frLIT.csv", sep=""))
str(lit.cities)
```


## PLOT FIG 2
```{r, echo=TRUE}
plot.new()
#svg("SVGfig.MAP.sightsXregionWlit.cities.svg", width=7, height=6)
plot(mx.shapes,col=alpha(all.states$col.names, 0.5), border = 'white') 
map.axes(cex.axis=0.75)

#plot city location where observations reported any time 1999-2015, use average lat/long for obs X city
points(avg.city.loc.MASTER$avg.cityLOG,avg.city.loc.MASTER$avg.cityLAT, #LONG. goes 1st
       col=alpha("black", 0.9), 
       bg=alpha("black", 0.5), pch=21, cex=1.1) 

#add extra cities not on citizen science data, but ARE in cited literature
points(lit.cities$long,lit.cities$lat, #LONG. goes 1st
       col=alpha("black", 0.9), 
       bg=alpha("black", 0.5), pch=21, cex=1.1)


#reg7.master #for legend
legdtxt <- reg7.master$reg7.cat
legdcol <- reg7.master$col.names

legend("bottomleft", # position
       legend = legdtxt, 
       title = "Regions",
       cex = 0.75,
       pch=21,
       pt.cex = 1.8,
       col = alpha(legdcol, 1), pt.bg = alpha(legdcol, 0.5),
       bty = "n",
       y.intersp=1.1)
#dev.off()

```



# FIGURE 3: dotplot of observations by city (grouped by region)

## prepare data
```{r, echo=TRUE}
#now add this new y-coord back to dcasted cit.sci data above and re-melt?
# use MELT.citiesXyrs.no0s from above, need all cit.sci data = TRUE match (don't need all MASTER cities to match, use that to plot axes/text)
MELT.citiesXyrs.no0s.wPLOTORD <- merge(MELT.citiesXyrs.no0s, MASTER.cities.CSciXcites.Wstates.Wreg7s, by=c("city", "state"), all.x=TRUE)

#check # records should be same -- looks good! both 175 records
str(MELT.citiesXyrs.no0s)
str(MELT.citiesXyrs.no0s.wPLOTORD) #use this now to plot dotplot (has new y coords to allow extra lines for citation-only reported cities)


#find unique x axis (years)
year.labs <- as.data.frame(unique(MELT.citiesXyrs.no0s.wPLOTORD$year))
names(year.labs) <- "year"
year.labs <- arrange(year.labs, year)
str(year.labs)
#unique(df[c("yad", "per")])
city.labs.new <- unique(MELT.citiesXyrs.no0s.wPLOTORD[c("yPlot.ord.CSciXcites", "city")])
city.labs.new <- arrange(city.labs.new, yPlot.ord.CSciXcites) #sort by y axis coords
length(MASTER.cities.CSciXcites.Wstates.Wreg7s$city)

#get all 7 regions with avg reg.order to add reg labels at right spots as axis labels
GRxreg.plotlabs <- group_by(MELT.citiesXyrs.no0s.wPLOTORD, reg7.cat, reg.order)
SUMxreg.plotlabs <- as.data.frame(summarise(GRxreg.plotlabs, avg.reg.plot.ord=mean(yPlot.ord.CSciXcites)))
SUMxreg.plotlabs <- arrange(SUMxreg.plotlabs, reg.order)
str(SUMxreg.plotlabs)

```

## PLOT DOTPLOT

```{r, echo=TRUE}
#START PLOT CITIES ON Y
plot.new()
#setwd("C:/Users/ehobs/Dropbox/P-monk MX invasion/MX-Monks/PLOTS/")
#svg("SVGfig.DOTPLOT.sightsXcityXregionWlegnd.Wcitations.svg", width=9, height=11.5)

par(mar=c(5,8,3,12))

plot(MELT.citiesXyrs.no0s.wPLOTORD$year, MELT.citiesXyrs.no0s.wPLOTORD$yPlot.ord.CSciXcites, type="p", axes=F, 
     ylim=rev(c(1,length(MASTER.cities.CSciXcites.Wstates.Wreg7s$city))), xlim=c(min(year.labs$year), max(year.labs$year)),
     xlab="Year", ylab=NA,
     pch=21, col=alpha(MELT.citiesXyrs.no0s.wPLOTORD$col.names, 0.9), bg=alpha(MELT.citiesXyrs.no0s.wPLOTORD$col.names, 0.4), 
     cex=log(MELT.citiesXyrs.no0s.wPLOTORD$num.reportsXcity)+0.5
)
#ADD LITERATURE POINTS
#points(x=1999, y=52, col="red")

#ADD AXES

axis(4, at=rev(1:length(MASTER.cities.CSciXcites.Wstates.Wreg7s$city)), 
     labels=rev(MASTER.cities.CSciXcites.Wstates.Wreg7s$city), las=2, cex.axis=0.6)
axis(1, at=seq(min(year.labs$year), max(year.labs$year)), 
     labels=seq(min(year.labs$year), max(year.labs$year), by=1),
     las=2, cex.axis=1)
#add region labels
axis(2, at=rev(c(SUMxreg.plotlabs$avg.reg.plot.ord)), labels=rev(SUMxreg.plotlabs$reg7.cat), 
     las=2, cex.axis=1, lwd.tick=0)

# add lines for start of ban
abline(v=2008, lty=1, col=alpha("grey", 0.6), lwd=2)
box()

#add text for citations
# use locXyrs.citations.Y

text(x=locXyrs.citations.wCOORDS$year.sighted, y=locXyrs.citations.wCOORDS$yPlot.ord.CSciXcites, 
     labels=locXyrs.citations.wCOORDS$cite.id, col="black", 
     cex=0.6)
#col=locXyrs.citations.wCOORDS$col.names,

#add first report (1994-1995 somewhere, in MX City) as 1998.5, then add note for real year
text(x=1998.5, y=62, 
     labels="A", col="black", 
     cex=0.6)

#add key for point sizes
legdtxt <- c("1", "10", "100")
legdptcex <- log(as.numeric(c("1", "10", "100")))+0.5

legend("bottomleft", # position
       legend = legdtxt, 
       title = "Number of reports",
       cex = 0.9,
       pch=21,
       pt.cex = legdptcex,
       col = alpha("grey", 0.9), pt.bg = alpha("grey", 0.6),
       bty = "n",
       y.intersp=1.1)

#add colored lines in margin for regions
GR.reg.maxmin.Y <- group_by(MELT.citiesXyrs.no0s.wPLOTORD, reg7.cat, reg.order)
SUM.reg.maxmin.Y <- as.data.frame(summarise(GR.reg.maxmin.Y, 
                                                  min.Y=min(yPlot.ord.CSciXcites)-0.3,
                                                  max.Y=max(yPlot.ord.CSciXcites)+0.3
                                                  ))
SUM.reg.maxmin.Y
SUM.reg.maxmin.Y$x <- 1998

#merge colors from master with point locations
SUM.reg.maxmin.Y <- merge(SUM.reg.maxmin.Y, select(reg7.master, reg7.cat, col.names), by="reg7.cat")
SUM.reg.maxmin.Y <- arrange(SUM.reg.maxmin.Y, reg.order)

#add rectangles
rect(xleft=1998.25-0.1, ybottom=SUM.reg.maxmin.Y$min.Y, 
     xright=1998.25+0.1, ytop=SUM.reg.maxmin.Y$max.Y,
     col = alpha(SUM.reg.maxmin.Y$col.names,0.4), border = alpha(SUM.reg.maxmin.Y$col.names,0.9), lty = 1, lwd = 1, xpd=TRUE)

#dev.off()

```
